Global multivariate model learning from hierarchically correlated data

### 2021-01-26 main.tex
first version intro - edits

\title{Global multivariate model learning from hierarchically correlated data}
\date{}

\author{Edwin Rodriguez Horta} 
\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
  Paris-Seine, Laboratoire de Biologie Computationnelle et
  Quantitative -- LCQB, Paris, France}
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Havana, Cuba}
%
\author{Pierre Barrat-Charlaix} 
\affiliation{Biozentrum, Universität Basel, Basel, Switzerland}
%
\author{Alejandro Lage} 
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Havana, Cuba}
%
\author{Martin Weigt} 
\email{Correspondence to: Martin Weigt, \bf{martin.weigt@upmc.fr}}
\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
  Paris-Seine, Laboratoire de Biologie Computationnelle et
  Quantitative -- LCQB, Paris, France}

@@

In the last decades, biology has seen impressive progress in experimental techniques which has resulted in a large increase of available data. 
This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences. [UNIPROT/GENBANK]
This has in turn fuelled a development of quantitative methods to model forces guiding the evolution of sequences or other biological traits [REFS?]. 
A common idea in this context is that it is possible to use characteristics of homologous genes or organisms to construct quantitative models of the selection acting on them. 
A successful example is the representation of protein sequences by probabilistic models in the so-called DCA method [REF].

Models built in this way usually assume that biological characteristics of related sequences or organisms are evolving under the same constraints but are independent from each other in the statistical sense. 
This allows simple formulation for model learning. 
However, phylogenetic relations between evolving entities imply that these cannot be considered independent from each other, and doing so can result in important biases when learning models. 
For instance, it has been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function [COLWELL paper].

In order to correct for these biases, it is necessary to disentangle the statistical effects due to selection forces and those due to phylogeny. 
However, this becomes a complex problem when dealing with complex models where traits under study are not independent from each other. 
Accounting for these biases then requires one to understand how correlated traits evolve along a tree. 
History of modeling this with BM or OU processes [Felsenstein, Hansen]. Also, field of PCM

Increasing data --> new quantitative techniques in biology. 
One big recent success: pairwise sequence models, blabla
However, models learned from data with phylogenetic correlations can suffer from biases. To account for these biases in the case of complex models, it's necessary to understand how correlated traits evolve along a tree (or: quantities that are modeled and typically not independent evolve along a tree)

There is a history of looking at multi parameter complex models and their relation with phylogeny. 
in this context, it is often assumed that quantitative traits that evolve under selection can be modeled using an OU process. 
This is usually done in the field of PCMs, with most common models being Brownian motion or OU processes.

Here, we want to understand how to find parameters values of an OU gaussian process from data correlated through a tree. 
This could be applied to the evolution of any correlated biological traits, but also potentially to protein sequences.

The MS is divided as follows: >>>

\paragraph{Old} % (fold)
\label{par:old}

Theory and applications of the inverse problem of statistical physics has had great development in recent years since the availability of large amounts of data on a microscopic scale produced by the continuous technological progress. In this context statistical physics paths: from model parameters to  observables, is reversed, toward find global models which describe the statistics of  a set of observed  configurations \cite{Inverse_problem_Berg}.

Most of the work has been focus in equilibrium reconstruction , where the aim is to learn max Entropy models, Boltzman distribution, from independent and identically distributed equilibrium configurations. Another kind of methods deal with parameters learning from either time series data or from samples of the non-equilibrium steady state: non-equilibrium reconstruction.

Less effort has been done for data biased by population structured signal, where  samples  are not independent because of presence of relatedness between configurations. These structure could be for instance of  geographical or demogrhapic  nature in a population or  phylogenetic  nature  related to a evolutionary processes shaping trait variation among species.  Here we are going to deal with data given by sequences phylogenetically related  through a tree, which describes the hierarchical correlation between configurations. An important problem is the  distinction   of patterns induced by species traits or by phylogeny.

Our source of inspiration came from the inverse problem in proteins \cite{Inverse_problem_proteins}, where  covariance signal present in an alignment of homologous proteins is described via Potts model. This covariance signal is assumed a consequence of phenotypics constraints like protein structure and function. However it is known that phylogeny relationships in homologous sequences  contaminate the statistics of  the alignment used to learn the equilibrium model. Therefore a direct inference on this alignments  leads to the existence of Potts parameters that attempt to model the full biased statistics limiting their capacity to predict phenotypic properties.

 To define and solve the inverse problem of statistical physics for data samples with hierarchical  relatedness structure is the aim of this work.

Main difficulties to address  this kind of problem  come from the necessity to use  a proper global evolutionary model which describes the relation between sequences edge connected in a tree and the discrete  characters of data variables which makes certain calculations intractable \cite{entropy_paper}.

Here we  adopt an approximation where we forget about the discrete nature of variables and model them by continuous variables instead. This transforms the Potts model into a Gaussian distribution \cite{Baldassi}, making the design of a global propagator tractable. We propose a model of evolution described by a multivariate Ornstein-Uhlenbeck process \cite{gardiner}, which is used  to generate evolutionary related data given an arbitrary phylogenetic tree. The corresponding inverse problem became in the inference of the Gaussian potential (Potts parameters) from the knowledge of tree leaves configurations and the dynamics of evolution.

To this aim, we first review in section \ref{sec:ornstein_uhlenbeck_dynamics}, main characteristics of a multivariate Ornstein-Uhlenbeck process. In section \ref{evolutionary process} we describe the evolutionary process, describing the algorithm used to generate data phylogenetically biased. The statement of the problem in a Bayesian framework is shown in section \ref{statement of the problem}. Section \ref{methods} is dedicated to describe methodology used to solve the inverse problem. Results for evolution of continuos variable sequences and  the application for configuration of discrete variables is presented in section \ref{Results}. Finally the conclusion and outlook in \ref{conc}.

% paragraph old (end)

@@

We consider a system characterized by $L$ continuous degrees of freedom and whose state is fully described by an $L$-dimensional vector $\vx\in\mathbb{R}^L$. These degrees of freedom can be continuous phenotypic traits of some living organism, or the sequence of a gene or a protein if a continuous approximation is made. 
At equilibrium, $\vx$ is assumed to be normally distributed,
\begin{equation}
	P_{eq}(\vx) = \frac{1}{Z(\bm{J})}\curlynormal{\vx^T\bm{J}\vx}\ ,
	\label{eq:eq_distribution}
\end{equation}
where $\bm{J}$ is the symmetric, positive definite \emph{coupling matrix} and $Z(\bm{J}) = \sqrt{(2\pi)^L/\det \bm{J}}$ is the normalization constant; the means of all components of $\vx$ are set to zero without loss of generality. 
We are interested in inferring the coupling matrix from a given amount of observed states $\vx$ of the system.
If these observations were independent from each other, due to the simple Gaussian form of Eq.~(\ref{eq:eq_distribution}), $\bm{J}$ would simply be equal to the inverse of the empirical \emph{covariance matrix} of the data, written $\bm{C}=\bm{J}^{-1}$.

However, here we are concerned with the case where observations are not independent. 
On the contrary, they result from a dynamical process taking place during a finite amount of time. 
Different datapoints are therefore correlated even if individually assumed to be at equilibrium. 
This dynamical process is described below.

We suppose that the considered system evolves according to the following Langevin equation 
\begin{equation}
	\gamma^{-1}\frac{\text{d}\vx}{\text{d}t} = - \bm{J}\vx + \vec{\xi}(t).
	\label{eq:langevin}
\end{equation}
Here, $\vec{\xi}(t)$ is a vector of uncorrelated white noise, and $\gamma^{-1}$ is the characteristic timescale governing the dynamics. 
In short, Eq.~(\ref{eq:langevin}) states that the system described by $\vx$ undergoes Brownian motion in a quadratic energy landscape characterized by the coupling matrix $\bm{J}$.

We are not interested in $\vx$ directly, but rather in its probability distribution $P(\vx\vert\,\vx_0,\Delta t)$, \ie~in the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ some time $\Delta t$ before. 
The Fokker-Planck equation corresponding to Eq.~(\ref{eq:langevin}) is straightforward to write, 
\begin{equation}
	\gamma^{-1} \partial_t P = \left(-\sum_{a,b=1}^L\frac{\partial}{\partial x_a} J_{ab}x_b + \sum_{a=1}^L\frac{\partial^2}{\partial x_a^2}\right) P, 
	\label{eq:FP}
\end{equation}
where the parenthesized expression on the right hand side is understood as an operator acting on $P$. 
The solution to Eq.~(\ref{eq:FP}) is a multivariate normal distribution~\cite{singh2017multiOU}:
\begin{equation}
	P(\vx|\, \vx_0, \Delta t) = \left[(2\pi)^N\det\Sig\right]^{-1/2}
	\curlynormal{(\vx-\vec{\mu})^T\iSig(\vx- \vec{\mu})},
	\label{eq:OUpropagator}
\end{equation}
where we introduce the matrices $\Sig$ and $\Lam$ as well as the vector $\vec{\mu}$ as 
\begin{equation}
	\Lam = e^{-\gamma\bm{J}}, \qquad \vec{\mu} = \Lam^{\Delta t} \vx_0, \qquad \Sig = \bm{J}^{-1}(\mathbb{1} - \Lam^{2 \Delta t}).
	\label{eq:def_lambda}
\end{equation}
Eqs.~(\ref{eq:OUpropagator}) and~(\ref{eq:def_lambda}) define a multivariate \emph{Ornstein-Uhlenbeck} (OU) process.

Note that since matrix $\Lam$ is an exponential of $\bm{J}$, it is symmetric, has strictly positive eigenvalues and commutes with $\bm{J}$. 
We also underline that $\Sig$ and $\vec{\mu}$ depend on $\Delta t$, although this dependence is not explicitly written in our notation to make it less heavy.

By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues, one immediately recovers Eq.~(\ref{eq:eq_distribution}). 
We can thus compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time difference $\Delta t$ by multiplying Eqs.~(\ref{eq:eq_distribution}) and~(\ref{eq:OUpropagator}), 
\begin{eqnarray}
    P(\vx_1, \vx_2\vert\,\Delta t) &= & P(\vx_1 |\, \vx_2,\Delta t) \times P_{eq}(\vx_2)
    \nonumber\\
    &\propto& \curlynormalpar{\vx_1^T\iSig\vx_1 + \vx_2^T\iSig\vx_2 - 2\vx_1^T\Lam^{\Delta t}\iSig\vx_2}. 
	\label{eq:OUjoint}	
\end{eqnarray}
This equation illustrates the \emph{time reversibility} of the OU process. 
Indeed, the distribution does not depend on which configuration $\vx_1$ or $\vx_2$ came first.

Equation \eqref{eq:OUjoint} allows for computing the joint covariance of the correlated equilibrium configurations $\vx_1$ and $\vx_2$. 
The probability distribution in Eq.~(\ref{eq:OUjoint}) is normal with an inverse covariance matrix defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
By inverting this block matrix, given that $\Lam$ and $\Sig$ commute and are invertible, one obtains the following covariance: 
\begin{equation}
	\langle \vx_1\vx_2^T \rangle_{\Delta t} = \Lam^{\Delta t}\bm{J}^{-1} = \Lam^{\Delta t}\bm{C}. 
	\label{eq:pairwisecov}
\end{equation}

Eq.~(\ref{eq:pairwisecov}) allows us to readily distinguish two regimes. 
Let us call $\rho_a$ the eigenvalues of $\bm{J}$. 
The eigenvalues of $\Lam^{\Delta t}\bm{C}$ are then equal to $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$. 
Since all $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially over time. The slowest timescale of exponential decay is set by $\tau_c^{-1} = \gamma\rho_{min}$, with $\rho_{min}$ being the smallest eigenvalue of $\bm{J}$. 
Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated. 
If this is verified for all pairs of observations $\vx_i$ and $\vx_j$, the regime is that of \emph{uncorrelated} data -- the inference of $\bm{J}$ can simply be performed by inverting the empirical covariance matrix extracted from the data. 
Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~(\ref{eq:pairwisecov}) becomes irrelevant. Actually, $\vx_1$ and $\vx_2$ coincide at that point, \ie~we have $P(\vx_1,\vx_2|\,\Delta t=0) = P_{eq}(\vx_1) \times \delta(\vx_1 - \vx_2)$ using the $L$-dimensional Dirac distribution.

@@

The problem discussed here is the inference of the probability distribution describing samples that are hierarchically correlated by a tree, cf.~Fig.~\ref{fig:sample_tree}. 
Formally, we assume that the data consists of $N$ real-valued vectors of length $L$, denoted $\{\vx_i\}\in\mathbb{R}^L$ with $i=1,...,N$. 
Taken individually, the $\vx_i$ are distributed according to Eq.~\ref{eq:eq_distribution}, \ie~according to a multivariate Gaussian of zero mean and covariance $\bm C$.
By construction, the equilibrium covariance between any pair of elements of a given vector $\vx=(x^1,...,x^L)$ is given by the inverse of the coupling matrix: $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$ for all $a,b=1,...,L$. 
This implies that inferring the coupling matrix defining the probability distribution amounts to finding the \emph{equilibrium} covariance matrix $\mathbf{C}$.

However, this covariance cannot be directly measured as we consider observations that are not independently distributed.
Instead, the set of measured configurations $\{\vx_i\}_{i=1,...,N}$ is the result of an Ornstein-Uhlenbeck (OU) process taking place on a tree $\mathcal{T}$, as is illustrated in Fig.~\ref{fig:sample_tree}:
\begin{itemize}
    \item The process starts at the root node $r$ with a state vector $\vx_r$ drawn from the equilibrium distribution $P_{eq}$.
    \item On each branch $(n,a(n))$ of length $\Delta t_{n,a(n)}$ connecting node $n$ with its ancestral node $a(n)$, the dynamics follow Eq.~(\ref{eq:langevin}), starting from initial condition $\vx_{a(n)}$, and running for time $\Delta t_{n,a(n)}$. In other words, given the state $\vx_{a(n)}$ of the ancestral node, $\vx_n$ is sampled from $P(\vx_n |\, \vx_{a(n)}, \Delta t_{n,a(n)})$, see Eq.~(\ref{eq:OUpropagator})
    \item As a consequence, OU processes on branches stemming from common ancestral node evolve independently, but from an identical initial condition.
    \item Observed data vectors correspond to the states of the leaves of the tree at the end of this process. The states of the internal nodes are not part of the observed data and remain unknown. 
\end{itemize}
This process is thought to represent the evolution of biologic traits along a phylogenetic tree, with the leave nodes corresponding to traits observed in today's species. Note that due to the reversible nature of our OU process, the joint probability of any pair of leaf configurations $\vx_i$ and $\vx_j$, with $i,j\in\{1,...,N\}$, is given by  $P(\vx_i,\vx_2|\,\Delta t_{ij})$ (Eq.~(\ref{eq:OUjoint}), with $\Delta t_{ij}$ denoting the total branch length of the path connecting $i$ and $j$ in the tree.

@@

We adopt a Bayesian inference approach by writing the probability of a given set of parameters $\{\bm C, \gamma\}$ given the data $\{\mathcal{D},\mathcal{T}\}$ using Bayes' equation
\begin{equation}
	P(\bm C, \gamma\vert\mathcal{D},\mathcal{T}) \propto P(\mathcal{D}\vert\bm C, \gamma,\mathcal{T})\cdot P(\bm C, \gamma),
	\label{eq:bayesian_likelihood}
\end{equation}
with the proportionality constant not depending on the parameters $\{\bm C, \gamma\}$. Here, $P(\bm C, \gamma)$ can be any arbitrarily chosen prior distribution. The difficulty in Eq.~(\ref{eq:bayesian_likelihood}) lies in the estimation of the likelihood $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$,
\ie~of the joint probability of the datapoints $\mathcal{D}=\{\vx_i\}_{i=1,...,N}$ for an OU process given by its parameters $\{\bm C, \gamma\}$ and the tree $\mathcal{T}$.  We detail the computation of this probability in the following section.

@@

The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by Eq.~(\ref{eq:OUjoint}) and corresponds to a joint normal distribution. 
This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with zero mean and variance described above in Eqs.~(\ref{eq:def_lambda}).
Of importance here is that this property of the OU process can be extended to the joint distribution of any subset of nodes in a tree. 
In other words, if we now define $\vec{X}=[\vx_1,\ldots,\vx_N]$ to be the concatenation of all configurations in our dataset $\mathcal D$, we can write the distribution of $\vec{X}$ as
\begin{equation}
	P(\vec{X}\vert\,\bm{C},\gamma,\mathcal{T}) = \left((2\pi)^{LN}\det\mathbb{G}\right)^{-\frac{1}{2}}\curlynormal{\vec{X}^T\mathbb{G}^{-1}\vec{X}},
	\label{eq:multiOU_jointall}
\end{equation}
where $\mathbb{G}$ is the \emph{joint covariance matrix} and depends on the tree as well as on $\bm{C}$ and $\gamma$.

>>>>

\title{Global multivariate model learning from hierarchically correlated data}
% \date{}

\author{Edwin Rodriguez Horta} 
\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
  Paris-Seine, Laboratoire de Biologie Computationnelle et
  Quantitative -- LCQB, Paris, France}
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Havana, Cuba}
%
\author{Alejandro Lage} 
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Havana, Cuba}
%
\author{Martin Weigt} 
\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
  Paris-Seine, Laboratoire de Biologie Computationnelle et
  Quantitative -- LCQB, Paris, France}
%

\author{Pierre Barrat-Charlaix} 
\email{Correspondence to: Pierre Barrat-Charlaix, \bf{pierre.barrat@unibas.ch}}
\affiliation{Biozentrum, Universit\"at Basel, Basel, Switzerland}

@@

In the last decades, biology has seen impressive progress in experimental techniques which has resulted in a large increase of available data. 
This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}. 
This has in turn fuelled a development of quantitative methods to model forces guiding the evolution of sequences or other biological traits [REFS/EXAMPLES?]. 
A common idea in this context is that it is possible to use characteristics of homologous genes or organisms to construct models of the selection acting on them. 
A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}.

Models built in this way usually assume that biological characteristics of related organisms are evolving under the same constraints but are independent from each other in the statistical sense. 
This allows simple formulation for model learning. 
However, the phylogenetic relations between evolving entities are in direct contradiction with this simplifying hypothesis, and ignoring them can result in biases when learning models. 
For instance, it has been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018}.

In order to correct for these biases, it is necessary to disentangle the effects due to selection forces and those due to phylogeny. 
However, this becomes a hard problem when dealing with complex models where traits under study are not independent from each other. 
Accounting for these biases then requires one to understand how interdependent traits under selection evolve along a tree. 
A historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models traits as a Gaussian vector evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
This method is commonly used in the field of phylogenetic comparative methods (PCM) \cite{bartoszek_phylogenetic_2012,mitov_fast_2020}.

This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,baldassi_fast_2014}. 
In this context, the equilibrium distribution reached by the OU process represents the probability distribution given by the DCA method, which can be used to predict non-trivial structural contacts in the protein fold, effects of amino-acid mutations or even designing novel functional sequences \cite{morcos_direct-coupling_2011, figliuzzi_coevolutionary_2016, russ_evolution-based_2020}.

In this work, we are interested in constructing an inference method for parameters of an OU process from data correlated through a tree.
Our approach is purely methodological, and the data can represent any set of continuous phenotypic traits, \emph{e.g.} from different organisms, with the tree indicating the phylogenetic relations between data points. 
Inferred parameters then represent the selection forces without biasing effects from the phylogeny. 
The manuscript is divided as follows: we first review in section \ref{sec:ornstein_uhlenbeck_dynamics} the main characteristics of the multivariate OU process. 
We then describe the setting of the inference problem that we want to solve in section \ref{sub:statement_of_the_problem}, propose a solution in sections \ref{sub:Calculation_of_the_likelihood} and \ref{sub:maximization_of_the_likelihood}.
Finally, we present results obtained on simulated data in section \ref{sec:Results}, with the context of pairwise models of protein sequences in mind.

@@

We consider a system characterized by $L$ continuous degrees of freedom and whose state is fully described by an $L$-dimensional vector $\vx\in\mathbb{R}^L$. These degrees of freedom can be continuous phenotypic traits of some living organism, or the sequence of a gene or a protein if a continuous approximation is made. 
At equilibrium, $\vx$ is assumed to be normally distributed,
\begin{equation}
	P_{eq}(\vx) = \frac{1}{Z(\bm{J})}\curlynormal{\vx^T\bm{J}\vx}\ ,
	\label{eq:eq_distribution}
\end{equation}
where $\bm{J}$ is the symmetric, positive definite \emph{coupling matrix} and $Z(\bm{J}) = \sqrt{(2\pi)^L/\det \bm{J}}$ is the normalization constant; the means of all components of $\vx$ are set to zero without loss of generality. 
We are interested in inferring the coupling matrix from a given amount of observed states $\vx$ of the system.
If these observations were independent from each other, due to the simple Gaussian form of Eq.~\eqref{eq:eq_distribution}, $\bm{J}$ would simply be equal to the inverse of the empirical \emph{covariance matrix} of the data, written $\bm{C}=\bm{J}^{-1}$.

However, we consider the case where observations are not independent. 
On the contrary, they result from a dynamical process taking place during a finite amount of time, and different data-points are therefore correlated to each other.
This dynamical process is described below.

We suppose that the considered system evolves according to the following Langevin equation 
\begin{equation}
	\gamma^{-1}\frac{\text{d}\vx}{\text{d}t} = - \bm{J}\vx + \vec{\xi}(t).
	\label{eq:langevin}
\end{equation}
Here, $\vec{\xi}(t)$ is a vector of uncorrelated white noise, and $\gamma^{-1}$ is the characteristic timescale governing the dynamics. 
In short, Eq.~\eqref{eq:langevin} states that the system described by $\vx$ undergoes Brownian motion in a quadratic energy landscape characterized by the coupling matrix $\bm{J}$.

We are not interested in $\vx$ directly, but rather in its probability distribution $P(\vx\vert\,\vx_0,\Delta t)$, \ie~in the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ some time $\Delta t$ in the past. 
The Fokker-Planck equation corresponding to Eq.~\eqref{eq:langevin} is straightforward to write, 
\begin{equation}
	\gamma^{-1} \partial_t P = \left(-\sum_{a,b=1}^L\frac{\partial}{\partial x_a} J_{ab}x_b + \sum_{a=1}^L\frac{\partial^2}{\partial x_a^2}\right) P, 
	\label{eq:FP}
\end{equation}
where the parenthesized expression on the right hand side is understood as an operator acting on $P$. 
The solution to Eq.~\eqref{eq:FP} is a multivariate normal distribution~\cite{singh2017multiOU}:
\begin{equation}
	P(\vx|\, \vx_0, \Delta t) = \left[(2\pi)^N\det\Sig\right]^{-1/2}
	\curlynormal{(\vx-\vec{\mu})^T\iSig(\vx- \vec{\mu})},
	\label{eq:OUpropagator}
\end{equation}
where we introduce the matrices $\Sig$ and $\Lam$ as well as the vector $\vec{\mu}$ as 
\begin{equation}
	\Lam = e^{-\gamma\bm{J}}, \qquad \vec{\mu} = \Lam^{\Delta t} \vx_0, \qquad \Sig = \bm{J}^{-1}(\mathbb{1} - \Lam^{2 \Delta t}).
	\label{eq:def_lambda}
\end{equation}
Eqs.~\eqref{eq:OUpropagator} and~\eqref{eq:def_lambda} define a multivariate \emph{Ornstein-Uhlenbeck} (OU) process.

Note that since matrix $\Lam$ is an exponential of $\bm{J}$, it is symmetric, has strictly positive eigenvalues and commutes with $\bm{J}$. 
We also underline that $\Sig$ and $\vec{\mu}$ depend on $\Delta t$, although this dependence is not explicitly written in our notation to make it less heavy. 
By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues, one immediately recovers Eq.~\eqref{eq:eq_distribution}, meaning that the OU process converges to the desired equilibrium distribution.

We can compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by a time $\Delta t$ by multiplying Eqs.~\eqref{eq:eq_distribution} and~\eqref{eq:OUpropagator}, 
\begin{eqnarray}
    P(\vx_1, \vx_2\vert\,\Delta t) &= & P(\vx_1 |\, \vx_2,\Delta t) \times P_{eq}(\vx_2)
    \nonumber\\
    &\propto& \curlynormalpar{\vx_1^T\iSig\vx_1 + \vx_2^T\iSig\vx_2 - 2\vx_1^T\Lam^{\Delta t}\iSig\vx_2}. 
	\label{eq:OUjoint}	
\end{eqnarray}
This equation illustrates the \emph{time reversibility} of the OU process. 
Indeed, the distribution is symmetric in $\vx_1$ or $\vx_2$ and does not depend on which configuration came first.

Equation \eqref{eq:OUjoint} allows for computing the joint covariance of the correlated equilibrium configurations $\vx_1$ and $\vx_2$. 
The probability distribution in Eq.~\eqref{eq:OUjoint} is normal with an inverse covariance matrix defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
By inverting this block matrix, given that $\Lam$ and $\Sig$ commute and are invertible, one obtains the following covariance: 
\begin{equation}
	\langle \vx_1\vx_2^T \rangle_{\Delta t} = \Lam^{\Delta t}\bm{J}^{-1} = \Lam^{\Delta t}\bm{C}. 
	\label{eq:pairwisecov}
\end{equation}

Eq.~\eqref{eq:pairwisecov} allows us to readily distinguish two regimes. 
Let us call $\rho_a$ the eigenvalues of $\bm{J}$. 
The eigenvalues of $\Lam^{\Delta t}\bm{C}$ are then equal to $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$. 
Since all $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially over time. The slowest timescale of exponential decay is set by $\tau_c^{-1} = \gamma\rho_{min}$, with $\rho_{min}$ being the smallest eigenvalue of $\bm{J}$. 
Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated. 
If this is verified for all pairs of observations $\vx_i$ and $\vx_j$, the regime is that of \emph{uncorrelated} data -- the inference of $\bm{J}$ can simply be performed by inverting the empirical covariance matrix extracted from the data. 
Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~\eqref{eq:pairwisecov} becomes irrelevant. Actually, $\vx_1$ and $\vx_2$ coincide at that point, \ie~we have $P(\vx_1,\vx_2|\,\Delta t=0) = P_{eq}(\vx_1) \times \delta(\vx_1 - \vx_2)$ using the $L$-dimensional Dirac distribution.

@@

The problem discussed here is the inference of the probability distribution describing samples that are hierarchically correlated by a tree, cf.~Fig.~\ref{fig:sample_tree}. 
Formally, we assume that the data consists of $N$ real-valued vectors of length $L$, denoted $\{\vx_i\}\in\mathbb{R}^L$ with $i=1,...,N$. 
Taken individually, we assume that the $\vx_i$ are distributed according to Eq.~\eqref{eq:eq_distribution}, \ie~according to a multivariate Gaussian of zero mean and covariance $\bm C$.
By construction, the equilibrium covariance between any pair of elements of a given vector $\vx=(x^1,...,x^L)$ is given by the inverse of the coupling matrix: $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$ for all $a,b=1,...,L$. 
This implies that inferring the coupling matrix defining the probability distribution amounts to finding the \emph{equilibrium} covariance matrix $\mathbf{C}$.

However, this covariance cannot be directly measured as we consider observations that are not independently distributed.
Instead, the set of measured configurations $\{\vx_i\}_{i=1,...,N}$ is the result of an Ornstein-Uhlenbeck (OU) process taking place on a tree $\mathcal{T}$, as is illustrated in Fig.~\ref{fig:sample_tree}:
\begin{itemize}
    \item The process starts at the root node $r$ with a state vector $\vx_r$ drawn from the equilibrium distribution $P_{eq}$.
    \item On each branch $(n,a(n))$ of length $\Delta t_{n,a(n)}$ connecting node $n$ with its ancestral node $a(n)$, the dynamics follow Eq.~\eqref{eq:langevin}, starting from initial condition $\vx_{a(n)}$, and running for time $\Delta t_{n,a(n)}$. In other words, given the state $\vx_{a(n)}$ of the ancestral node, $\vx_n$ is sampled from $P(\vx_n |\, \vx_{a(n)}, \Delta t_{n,a(n)})$, see Eq.~\eqref{eq:OUpropagator}
    \item As a consequence, OU processes on branches stemming from common ancestral node evolve independently, but from an identical initial condition.
    \item Observed data vectors correspond to the states of the leaves of the tree at the end of this process. The states of the internal nodes are not part of the observed data and remain unknown. 
\end{itemize}
This process is thought to represent the evolution of biologic traits along a phylogenetic tree, with the leave nodes corresponding to traits observed in today's species. Note that due to the reversible nature of our OU process, the joint probability of any pair of leaf configurations $\vx_i$ and $\vx_j$, with $i,j\in\{1,...,N\}$, is given by  $P(\vx_i,\vx_j|\,\Delta t_{ij})$ (Eq.~\eqref{eq:OUjoint}), with $\Delta t_{ij}$ denoting the total branch length of the path connecting $i$ and $j$ in the tree.

@@

We adopt a Bayesian inference approach by writing the probability of a given set of parameters $\{\bm C, \gamma\}$ given the data $\{\mathcal{D},\mathcal{T}\}$ using Bayes' equation
\begin{equation}
	P(\bm C, \gamma\vert\mathcal{D},\mathcal{T}) \propto P(\mathcal{D}\vert\bm C, \gamma,\mathcal{T})\cdot P(\bm C, \gamma),
	\label{eq:bayesian_likelihood}
\end{equation}
with the proportionality constant not depending on the parameters $\{\bm C, \gamma\}$. Here, $P(\bm C, \gamma)$ can be any arbitrarily chosen prior distribution. The difficulty in Eq.~\eqref{eq:bayesian_likelihood} lies in the estimation of the likelihood $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$,
\ie~of the joint probability of the datapoints $\mathcal{D}=\{\vx_i\}_{i=1,...,N}$ for an OU process given by its parameters $\{\bm C, \gamma\}$ and the tree $\mathcal{T}$.  We detail the computation of this probability in the following section.

@@

The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by Eq.~\eqref{eq:OUjoint} and corresponds to a joint normal distribution. 
This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with zero mean and variance described above in Eqs.~\eqref{eq:def_lambda}.
Of importance here is that this property of the OU process can be extended to the joint distribution of any subset of nodes in a tree. 
In other words, if we now define $\vec{X}=[\vx_1,\ldots,\vx_N]$ to be the concatenation of all configurations in our dataset $\mathcal D$, we can write the distribution of $\vec{X}$ as
\begin{equation}
	P(\vec{X}\vert\,\bm{C},\gamma,\mathcal{T}) = \left((2\pi)^{LN}\det\mathbb{G}\right)^{-\frac{1}{2}}\curlynormal{\vec{X}^T\mathbb{G}^{-1}\vec{X}},
	\label{eq:multiOU_jointall}
\end{equation}
where $\mathbb{G}$ is the \emph{joint covariance matrix} and depends on the tree as well as on $\bm{C}$ and $\gamma$.
